---
title: "Image Segmentation"
author: "Kristin Linn"
output:
  ioslides_presentation:
    widescreen: yes
    css: ../styles.css
bibliography: ../refs.bib       
---

```{r setup, include=FALSE}
library(methods)
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

## Loading Data
- Let's read in the training T1s and brain mask for subject 05.

```{r loading}
library(ms.lesion)
library(neurobase)
library(fslr)
library(scales)
library(extrantsr)
files = get_image_filenames_list_by_subject(
  group = "training", 
  type = "coregistered")
t1s = lapply(files, function(x) readnii(x["MPRAGE"]))
masks = lapply(files, function(x) readnii(x["Brain_Mask"]))
```


## MS Lesion Segmentation with OASIS
- OASIS is Automated Statistical Inference for Segmentation [@sweeney2013oasis]
- The OASIS algorithm takes FLAIR, T1, T2, and PD images from patients with multiple sclerosis (MS) and produces OASIS probability maps of MS lesion presence, which can be thresholded into a binary lesion segmentation.
- The OASIS library in R provides default parameters from a previously trained data set.
- To retrain the model using new data, binary masks of `gold standard’ lesion segmentations are needed and should be in T1 space.

## MS Lesion Segmentation with OASIS
- Let's read in the other modalities and manual segmentations

```{r readnii_show_run}
t2s = lapply(files, function(x) readnii(x["T2"]))
flairs = lapply(files, function(x) readnii(x["FLAIR"]))
pds = lapply(files, function(x) readnii(x["PD"]))
golds = lapply(files, function(x) readnii(x["mask2"]))
```

## MS Lesion Segmentation with OASIS
- Here's the T1 volume for training subject 05 with the ‘gold standard’ manual lesion segmentation overlayed.

```{r over_show_run}
les_mask = golds$training05
ortho2(t1s$training05, les_mask, col.y = "orange")
```

## Making OASIS data frames

- OASIS requires a particular data frame format
- OASIS has an option to preprocess your data for you (preproc)
- OASIS has an option to normalize the intensities of your data for you using a whole-brain normalization (normalize)
- `make_df()` below is a helper function

```{r oasis_df_show, eval=FALSE}
library(oasis)
make_df = function(x){
  res = oasis_train_dataframe(
      flair=flairs[[x]], t1=t1s[[x]], 
      t2=t2s[[x]], pd=pds[[x]], 
      gold_standard=golds[[x]], 
      brain_mask=masks[[x]], 
      preproc=FALSE, normalize=TRUE, 
      return_preproc=FALSE)
  return(res$oasis_dataframe)
}
oasis_dfs = lapply(1:5, make_df)
```

## Training OASIS

```{r oasis_run_show, eval=FALSE}
oasis_model = do.call("oasis_training", args = oasis_dfs)
```


## References {.smaller}
