---
title: "Building Distributable Packages"
author: "John Muschelli"
date: "March 28, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

- Naming multiple functions
- Package Vignettes
- Rbuildignore
- Including "data" in a package
- Installing from GitHub
- Checking with Travis CI/Appveyor/Coveralls
- Unit Testing
- Building a source tarball

## Quick naming convention guidelines

These are personal but serve me well:

- All of my functions are lowercase with underscores separating words
- The name of the file is `function_name.R` where the name of the function is `function_name`.  This helps me find things
    - This means I have a **lot** of files in R/ folder.
    - I use the go to file/function search to easily search through them
- If you have multiple functions that should have the same help file, you can use the `@rdname` tag, making sure you still put `@export` at the second function definition.
- Don't make filenames too long (over 25-30 characters max)
- Never put spaces in filenames.

## Package Vignettes

Vignettes are package tutorials to show how the package should work

- start with typing `devtools::use_vignette("My_Vignette")`.  Do NOT put spaces in the name.  It's allowable, but just don't.
- Change the title and VignetteEntry tag.
- Write vignette using RMarkdown (making sure to use `library` calls)
    - make sure all packags in `library` calls are in at LEAST Suggests, if not Imports/Depends
    
## Rbuildignore

The file `.Rbuildignore` has a period in front of it.  That means it's hidden (depending on your settings for folders).  

Go to go to file/function search bar and type `.Rbuildignore`.  Anything in here says "when you go to build R package, don't worry about these files".  Wildcards are fine (like `*`).  Use escapes `\` to escape periods because they are wildcards.

For example, if you have a `README.md` file that is for GitHub but not your package:
```
README\.md$
```
would be in `.Rbuildignore`

## Including data

See http://r-pkgs.had.co.nz/data.html first. From there:

> The most common location for package data is (surprise!) data/. 
> Each file in this directory should be a .RData file created by save() containing a single object (with the same name as the file).

I name mine with `.rda`.  Either use `.rda` or `.RData` - NO OTHER SPELLINGS - it's important.

```{r, eval = FALSE}
blah = matrix(rnorm(100), nrow = 5)
save(blah, file = "data/blah.rda", compression_level = 9)
```

## Documenting data: in `R/blah.R`

```
#' @title Random data, very blah.
#'
#' @description A dataset I made as an illustration.
#'
#' @format A matrix with 20 rows and 5 columns:
#' \describe{
#'   \item{col1}{random col}
#'   \item{col2}{random col}
#'   ...
#' }
"blah"
```

## Documenting "data"

Data here is not always a `data.frame` or `matrix`.  It can be a `lm` or `glm` object.  

You can make sure you reference these specifically by:

```{r, eval = FALSE}
mypkg::blah
```

anywhere in the code.

There are ways to include non-rda files using `inst`, but we won't cover that here



## Installing from GitHub

If you have pushed your package to GitHub, you can install it using:

```{r, eval = FALSE}
devtools::install_github("YOURUSERNAME/mypkg")
```
where `mypkg` is the name of the repo.  Make the repo name the same name as the package. It's just easier.  This may be replaced in the future by:
```{r, eval = FALSE}
remotes::install_github("YOURUSERNAME/mypkg")
```

## Checking with Travis CI/Appveyor/Coveralls

`use_travis`, `use_appveyor`, `use_coverage`

These will add YAML files to your packages on Github for testing in Linux/Mac (Travis), Windows (Appveyor), and Code coverage (Coveralls/Travis).

## Version Numbering

From http://r-pkgs.had.co.nz/release.html#release-version use `major.minor.patch.dev`:

* Increment patch, e.g. 0.8.2 for a patch: you’ve fixed bugs without adding any significant new features. 

* Increment minor, e.g. 0.9.0, for a minor release. A minor release can include bug fixes, new features and changes in backward compatibility. 

* Increment major, e.g. 1.0.0, for a major release. This is best reserved for changes that are not backward compatible and that are likely to affect many users. 


## Unit Testing

`use_testthat` - use the `testthat` package to test a package

- if you change function A, does it work with the cases it used to?  There should be a test for that
    - 







