---
title: "Local R Setup"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true   
    theme: cosmo
  beamer_presentation: default
---


```{r, echo = FALSE}
library(methods)
library(knitr)
opts_chunk$set(eval = FALSE)
```


# System Requirements

The functionality of some of these packages requires Linux, Mac OSX, or a Unix-type system.   This excludes Windows in some cases.  The code was run on Linux and Mac OSX.  Although we do not currently support it, [Windows 10 has a Linux subsystem ](https://msdn.microsoft.com/en-us/commandline/wsl/install_guide), which may allow for a working installation of all the dependencies of Neuroconductor. 

Although still untested there is a good [tutorial on running FSL on Windows](http://www.nemotos.net/?p=1481) as well as [ANTsR on Windows](https://github.com/stnava/ANTsR/wiki/Installing-ANTsR-in-Windows-10-(along-with-FSL,-Rstudio,-Freesurfer,-etc)).  

## Install R

You can install the latest R from [http://cran.r-project.org/](http://cran.r-project.org/).

We also recommend installing the  [RStudio IDE](https://www.rstudio.com/products/rstudio/download/)

## CMake and Git

These are required by the `ANTsR` suite of tools.  To install these, see the relevant sections in the [Installing ANTsR tutorial ](http://johnmuschelli.com/neuroc/installing_ANTsR/index.html)


## FSL
See the installation instructions for FSL from [https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FslInstallation](https://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FslInstallation).  

### Neurodebian
The following was done to make the Neurodebian installation look like that of Mac OSX:
```bash
##########################################
# FSL
##########################################
sudo apt-get update \
    && sudo apt-get install -y fsl-complete

# Debian has a fixed FSLDIR
export FSLDIR=/usr/local/fsl
export FSLSHARE=/usr/share/data
# add the fsl binary path to the search path
export PATH=$PATH:${FSLDIR}/bin
# export PATH=$PATH:/usr/lib/fsl/5.0
# Possum is installed in the same directory
export POSSUMDIR=$FSLDIR

sudo mkdir -p ${FSLDIR}/bin
sudo cp /usr/lib/fsl/5.0/* ${FSLDIR}/bin/
sudo mkdir -p ${FSLDIR}/lib
sudo mv ${FSLDIR}/bin/lib* ${FSLDIR}/lib/

export FSLOUTPUTTYPE=NIFTI_GZ
# export LD_LIBRARY_PATH=$LD_LIBRARY_PATH:/usr/lib/fsl/5.0
export LD_LIBRARY_PATH=${FSLDIR}/lib/:$LD_LIBRARY_PATH:

# sudo cp /usr/share/fsl/5.0/etc/fslconf/fsl.sh $FSLDIR/etc/fslconf/fsl.sh
export FSLOUTPUTTYPE=NIFTI_GZ
# sudo rm $FSLDIR/etc/fslconf/fsl.sh
sudo mkdir -p $FSLDIR/etc/fslconf/
fname=${FSLDIR}/etc/fslconf/fsl.sh
echo "FSLDIR=/usr/local/fsl" > ${fname}
echo 'PATH=${PATH}:${FSLDIR}/bin' >> ${fname}
echo 'LD_LIBRARY_PATH=${FSLDIR}/lib:$LD_LIBRARY_PATH' >> ${fname}
echo "FSLOUTPUTTYPE=NIFTI_GZ" >> ${fname}

##########################
# Setting global library
##########################
sudo echo "${FSLDIR}/lib" > /etc/ld.so.conf.d/fsl.conf
sudo ldconfig

sudo mkdir -p ${FSLDIR}/data/standard
sudo mkdir -p ${FSLDIR}/data/atlases 

#######################################
# Setting things up like other installers
#######################################
# settring up standard
sudo cp -R ${FSLSHARE}/fsl-mni152-templates/* ${FSLDIR}/data/standard/

# setting up atlases
sudo cp -R ${FSLSHARE}/harvard-oxford-atlases/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/juelich-histological-atlas/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/bangor-cerebellar-atlas/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/jhu-dti-whitematter-atlas/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/forstmann-subthalamic-nucleus-atlas/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/fsl-resting-connectivity-parcellation-atlases/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/mni-structural-atlas/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/oxford-thalamic-connectivity-atlas/* ${FSLDIR}/data/atlases/
sudo cp -R ${FSLSHARE}/talairach-daemon-atlas/* ${FSLDIR}/data/atlases/ 
```

# R Packages
## Devtools Package

The tutorial and breakdown for installing the `devtools` package is located at: [https://www.rstudio.com/products/rpackages/devtools/](https://www.rstudio.com/products/rpackages/devtools/).  Please read through it as most of the code below requires `devtools`.

## Bioconductor
We need Bioconductor packages in some cases and we can install it with the code below:

```{r}
# Installing Bioconductor
library(utils); 
source("http://bioconductor.org/biocLite.R"); 
biocLite(pkgs = c("Biobase"), suppressUpdates = TRUE, suppressAutoUpdate = TRUE, ask = FALSE)
```

## RGL

## Mac and Linux
The `rgl` package cand be installed using `install.packages` usually.

## Linux
The `install.packages` command may have some problems on Linux machines.  It can be installed in `bash` using: 
```bash
sudo apt-get -y update && sudo apt-get build-dep -y r-cran-rgl 
```

There may be an error that may be mitigated by the following command in bash:
```bash
sudo sed -i -- 's/#deb-src/deb-src/g' /etc/apt/sources.list && sudo sed -i -- 's/# deb-src/deb-src/g' /etc/apt/sources.list
```
and re-running the code above.


## Neuroimaging packages on Github

### Standard Packages
```{r}
source("https://neuroconductor.org/neurocLite.R")
# DICOM converter
neuro_install("dcm2niir")
library(dcm2niir); install_dcm2nii()
# HCP database connector
neuro_install("neurohcp")
# dcm2nii Rcpp implementation
neuro_install(c("RNifti", "divest", "oro.dicom", "oro.nifti", "WhiteStripe"))
# data for whitestripe
library(WhiteStripe); download_img_data()

# neurobase package
neuro_install("neurobase")
neuro_install("fslr")
```


<!-- source("https://neuroconductor.org/neurocLite.R") -->
<!-- # DICOM converter -->
<!-- devtools::install_github("muschellij2/dcm2niir") -->
<!-- library(dcm2niir); install_dcm2nii() -->
<!-- # HCP database connector -->
<!-- devtools::install_github("muschellij2/neurohcp") -->
<!-- # dcm2nii Rcpp implementation -->
<!-- install.packages(c("RNifti", "divest", "oro.dicom", "oro.nifti", "WhiteStripe")) -->
<!-- # data for whitestripe -->
<!-- library(WhiteStripe); download_img_data() -->

<!-- # neurobase package -->
<!-- devtools::install_github("muschellij2/neurobase") -->
<!-- devtools::install_github("muschellij2/fslr") -->


### ANTsR

Please see the the [Installing ANTsR tutorial ](http://johnmuschelli.com/neuroc/installing_ANTsR/index.html) if anything errors here:

```{r}
##########################################
# ANTs
##########################################
neuro_install("ITKR")
neuro_install("ANTsRCore", upgrade_dependencies = FALSE)
neuro_install("ANTsR", upgrade_dependencies = FALSE)
neuro_install("extrantsr", upgrade_dependencies = FALSE)
```

### Knitting using RStudio
```{r}
install.packages(c("formatR", "caTools", "rprojroot", "rmarkdown"))
```

### Imaging Packages and Data

The only package that isn't directly available is the `ms.lesion` package, which is due to restrictions on the data.  This should be available from the tutorial

```{r}
##############################
# Install MS LESION DATA!
# INSTALL KIRBY21
##############################
neuro_install("papayar")
neuro_install("oasis")
neuro_install("malf.templates")
neuro_install("kirby21.t1")
```

# Conclusion
If all of these install correctly, restart `R` and test:

## Working installation of FSL and fslr test

```{r}
library(fslr)
have_fsl()
example("fsl_smooth")
```

## Working installation of ANTsR and extrantsr

```{r}
library(extrantsr)
example("smooth_image")
```
