---
title: "R Tutorial"
author: "Kristin Linn"
output:
  ioslides_presentation:
    widescreen: yes
    css: ../styles.css
bibliography: ../refs.bib       
---

```{r setup, include=FALSE}
library(methods)
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

## Help
- To see the documentation for a function, use the ? symbol before the name of the function.
- For example, to see documentation for `solve`:

```{r help}
?solve
```
- To search for help files, use a double ??

```{r help2}
??solve
```

## Some Details
- R is case sensitive
- Commands separated by new line or by a colon (;)
- Use # to comment
- Functions and commands in a separate file, say myCode.R, may be executed using 

```{r source, eval=FALSE}
source(myCode.R)
```

## Objects
- `ls()' or `objects()' may be used to display names of objects that are currently stored in the R session.
- Objects can be variables, arrays, character strings, functions, etc.
- Assignment by `<-' or `='.

## Other stuff
- Need to explain types, subsetting, and do.call
- Need to explain file.path and system.? commands?






