---
title: "R Tutorial"
author: "Kristin Linn"
output:
  ioslides_presentation:
    widescreen: yes
    css: ../styles.css
bibliography: ../refs.bib       
---

```{r setup, include=FALSE}
library(methods)
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

## Help
- To see the documentation for a function, use the ? symbol before the name of the function.
- For example, to see documentation for `solve`:

```{r help}
?solve
```
- To search for help files, use a double ??

```{r help2}
??solve
```

## Some Details
- R is case sensitive
- Commands separated by new line or by a colon (;)
- Use # to comment
- Functions and commands in a separate file, say myCode.R, may be executed using 

```{r source, eval=FALSE}
source(myCode.R)
```

## Objects
- `ls()` or `objects()` may be used to display names of objects that are currently stored in the R session.
- Objects can be variables, arrays, character strings, functions, etc.
- Assignment by `<-` or `=`.

## Vectors
- Create a vector of numeric values and assign to variable `v`
```{r vec}
v = c(1, 4, 3, 7, 8)
print(v)
```
- Subsetting:
```{r vecSub}
print(v[4])
print(v[1:3])
print(v[c(1,3,5)])
```

## Matrices
- Create a 3 x 4 numeric matrix and assign to variable `m`
```{r mat}
m = matrix(1:12, nrow=3)
print(m)
```
- Subsetting:
```{r matSub}
print(m[,4])
print(m[2,])
print(m[1,3])
print(m[1:2,3:4])
```
## Lists
- Initialize an empty list and add two elements to it
```{r list}
l = list()
l[[1]] = v
l[[2]] = m
print(l)
```
- Subsetting:
```{r listSub}
print(l[[1]])
```

## Writing functions
```{r fun}
square = function(x){
    return(x^2)
}
print(square(5))
print(square(c(1, 10, 100)))
```

## Loops and if statements
```{r loop}
for(i in 1:3){
  for(j in 2:4){
    if(j >= i){
      x = j - i
      print(x)
    } else{
    print("negative")
    }
  }
}
```

## Vectorization
- apply() function

## Other stuff
- Need to explain types, subsetting, and do.call
- Need to explain file.path and system.? commands?






